### 2016-07-06 case-studies.tex
Latest comments by KC and Kia addressed.

\paragraph{A video view counter.} Our first application is a Rails
implementation of a counter to count the number of views a video has
garnered on a video-sharing website like YouTube. The application
stores the video count in a MySQL database, and supports a read
operation (a database read transaction), and an increment operation (a
read-increment-write sequence wrapped in a transaction).  The
application is expected to preserve a \emph{monotonicity invariant},
namely that the view count should never appear to be decreasing.
However, the default \iso{Repeatable Read} isolation level of MySQL
leads to anomalous executions - with 32 concurrent writers performing
single increments, and one reader issuing a constant stream of reads,
we observed 12 violations of monotonicity in 10 rounds. We
subsequently instantiated our reasoning framework for an {\sc sc}
store with ANSI SQL isolation levels, and spent 5 man hours
determining appropriate isolation levels needed to enforce the
invariant. In particular, we were able to formally prove that to
preserve the invariant, writers need to execute using
\iso{Serializable} isolation, while readers can execute using
\iso{Read Uncommitted}. Repeating the experiments with this
configuration led to no violations, as expected. While executing all
transactions at {\sc ser} also prevents violations, it led to an
additional 113\% increase in the latency of writes averaged over 5
rounds with 32 concurrent writers.  We repeated the experiments with
the same configuration on Postgres, and found no violations. However,
Postgres' isolation levels~\cite{postgresiso} are stronger than ANSI
SQL's in non-trivial ways. We therefore formalized Postgres'
guarantees and repeated relevant parts of the proof to discover that
\iso{Repeatable Read} is in fact sufficient for writers - an
observation further validated through experiments. As usual, using
{\sc ser} for all transactions also suffices, but on Postgres results
in an additional 47\% increase in write latency.

>>>>

\paragraph{A video view counter.} Our first application is a Rails
implementation of a counter to count the number of views a video has
garnered on a video-sharing website like YouTube. The application
stores the video count in a MySQL database, and supports a read
operation (a database read transaction), and an increment operation (a
read-increment-write sequence wrapped in a transaction).  The
application is expected to preserve a \emph{monotonicity invariant},
namely that the view count should never appear to be decreasing.
However, the default \iso{Repeatable Read} isolation level of MySQL
leads to anomalous executions - with 32 concurrent writers performing
single increments, and one reader issuing a constant stream of reads,
we observed 12 violations of monotonicity in 10 rounds. We
subsequently instantiated our reasoning framework for an {\sc sc}
store with ANSI SQL isolation levels, and spent 5 man hours
determining appropriate isolation levels needed to enforce the
invariant. In particular, we were able to formally prove that to
preserve the invariant, writers need to execute using
\iso{Serializable} isolation, while readers can execute using
\iso{Read Uncommitted}. Repeating the experiments with this
configuration led to no violations, as expected. While executing all
transactions at {\sc ser} also prevents violations, it led to an
additional 113\% increase in the latency of writes averaged over 5
rounds with 32 concurrent writers.  We repeated the experiments with
the same configuration on Postgres, and found no violations. However,
Postgres' isolation levels~\cite{postgresiso} are stronger than ANSI
SQL's in non-trivial ways. We therefore formalized Postgres'
guarantees and repeated relevant parts of the proof to discover that
Postgres' \iso{Repeatable Read} is in fact sufficient for writers - an
observation further validated through experiments. As usual, using
{\sc ser} for all transactions also suffices, but on Postgres results
in an additional 47\% increase in write latency.

### 2016-07-06 motivation.tex
Latest comments by KC and Kia addressed.

As an alternative, consider the execution of this transaction under a
\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}.\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} that read uncommitted
data. In the current example, {\sc rc} isolation admits the two
executions shown in Fig.~\ref{fig:rc-ex} on a strongly consistent
({\sc sc}) store, such as a conventional RDBMS.

@@

\begin{figure}
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
\caption{\small Visibility and interference properties for different weak isolation levels can be
  captured by different instantiations of their happens-before relation.}
\label{fig:interference-ex}
\end{figure}

@@

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the definition of the
\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
what actions become visible to a transaction, or its constituents, and
when. To do so, we axiomatize the $\hbZ$ relation to capture the
interference characteristics of different isolation levels as their
specifications. Two such specifications are shown\footnote{Note that
specifications presented here are solely for illustration. The actual
specifications described in \S\ref{sec:ansi-isolation} are more
nuanced.} in Fig.~\ref{fig:interference-ex}. The \iso{Read Committed}
specification ($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to
happen-before an operation $\eta_2$ of $T_2$ only if every operation
in $T_1$ (including its commit) happens-before $\eta_2$ (we let $T_1
\hboar \eta_2$ denote this). The specification does not require $T_1$
to execute and commit before all operations of $T_2$, thus allowing it
to interfere in $T_2$. The \iso{Snapshot Isolation} specification
($\psi_{SI}$) requires two transactions that write to the same
variable (denoted $\wrstoar$)to be related by happens-before.  This
effectively prohibits interference due to actions in $T_1$ being
interleaved in $T_2$ (or, vise versa).

@@

We have thus far assumed a strongly consistent ({\sc sc}) store that,
in the absence of transactions with special isolation requirements,
makes the effects of any operation immediately visible to all
subsequent operations. The natural behaviour of {\sc sc} to totally
order all operations w.r.t. the $\hbZ$ relation could be in conflict
with the constraints imposed by weak isolation. To meet the
requirements of weaker isolation levels, say {\sc rc}, the store has
to adapt itself to \emph{hide} the effects of concurrent transactions
until they are committed. Once committed though, effects need to be
immediately visible to subsequent operations. This semantics can be
built into the reasoning framework, leading to a proof system
tailor-made for such stores. However, for the reasoning framework to
be truly useful, it has to be parametric over different consistency
semantics, supporting stores that are weaker than {\sc sc} (e.g.,
causally or eventually consistent), and should be able to reconcile
conflicts between consistency and isolation constraints.  We
demonstrate how our reasoning framework makes this possible in the
following sections.

>>>>

As an alternative, consider the execution of this transaction under a
\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}.\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} of uncommitted data. In
the current example, {\sc rc} isolation admits the two executions
shown in Fig.~\ref{fig:rc-ex} on a strongly consistent ({\sc sc})
store, such as a conventional RDBMS.

@@

\begin{figure}
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
\caption{\small Interference properties for different weak isolation
levels can be captured as constraints over the happens-before
relation.}
\label{fig:interference-ex}
\end{figure}

@@

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the nature of the
\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
what other transactions become visible to a transaction, or its
constituents, and when. To do so, we axiomatize the $\hbZ$ relation to
capture the interference characteristics of different isolation levels
as their specifications. Two such specifications are
shown\footnote{Note that specifications presented here are solely for
illustration. The actual specifications described in
\S\ref{sec:ansi-isolation} are more nuanced.} in
Fig.~\ref{fig:interference-ex}. The \iso{Read Committed} specification
($\psi_{RC}$) allows an operation $\eta_1$ of a transaction $T_1$ to
happen-before an operation $\eta_2$ of another transaction $T_2$ only
if every operation in $T_1$ (including its commit) happens-before
$\eta_2$ (we let $T_1 \hboar \eta_2$ denote this). The specification
does not require $T_1$ to execute and commit before all operations of
$T_2$, thus allowing it to interfere in $T_2$, just as \C{Wd2}
interferes in \C{Wd1} in Fig.~\ref{fig:motiv-eg-1-a}. The
\iso{Snapshot Isolation} specification ($\psi_{SI}$) requires two
transactions that write to the same variable (denoted $\wrstoar$) to be
related by happens-before.  This effectively prohibits interference
due to actions in $T_1$ being interleaved in $T_2$ (or, vice versa).

@@

Implicit in the above discussion is the assumption of a strongly
consistent ({\sc sc}) store that guarantees the visibility of all
previously committed transactions. {\sc sc} semantics can be built
into the reasoning framework, leading to a proof system tailor-made
for such stores. However, for the reasoning framework to be truly
useful, it has to be parametric over different consistency semantics,
supporting stores that are weaker than {\sc sc} (e.g., causally or
eventually consistent), and should be able to reconcile conflicts
between consistency and isolation constraints.  We demonstrate how our
reasoning framework makes this possible in the following sections.

### 2016-07-06 store-consistency.tex
Latest comments by KC and Kia addressed.

First, we observe that the consistency level of a data store can be
captured by store-specific consistency constraints, along with
transaction-specific isolation constraints, via the trace invariant
$\I$. In particular, we can split $\I$ into two components: (1).
$\I_s$, the store-specific invariant, and (2). $\I_c$, the
program-specific (or, client-specific) invariant, to capture
consistency and isolation constraints, respectively.  $\I$ is now a
conjunction: $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often
simplified to $\I \,=\, \I_s\wedge\I_c$).  While the program-specific
trace invariant ($\I_c$) is independent of the underlying data store,
the store specific invariant ($\I_s$) changes from store to store
depending on the consistency level.  For an eventually consistent
store, $\I_s$ is simply \emph{true}. Stronger stores, such as those
that support strong consistency, have non-trivial definitions for
$\I_s$.

@@

Notably, our view of replication does not explicitly involve replicas.
Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
weakly-consistent replication with a conventional replica-based model.
Under our model, the notion of a replica is subsumed by the concept of
visibility; a replica is defined by the subset ($S$) of global state
($\E.\A$) that an operation witnesses. Constraints over replica states
therefore manifest as constraints over a specific visibility relation.
For example, instead of requiring the store to be causally consistent,
an operation can witness a causally consistent subset of the state;
such demands can be made via the trace invariant $\I$. For a
pre-condition ($P$) of the operation to be useful, it has to be an
assertion over every causally consistent subset of the global state.
Since any replica that eventually executes the operation has to expose
one such subset ($S$), the pre-condition is guaranteed to hold
regardless of the replica. There is however one problem with this
explanation - by considering subsets of just one global state, it
ignores the fact that the global state (hence, the replica states)
change during the execution of the operation. To account for such
changes, we might choose to distinguish between an effect generation
event at one replica $r_1$ and an effect merge event at replica $r_2$,
requiring that \emph{non-conflicting} operations execute between these
two events at $r_2$, and that they preserve certain
invariants~\cite{gotsmanpopl16}.  Instead, our framework folds all such
machinery into a stability condition predicated on $\I$
(\S\ref{sec:rely-guarantee}).  Since any change to the global state
during the execution of the operation is an interference, and $P$ is
required to be stable with respect to any such interference, it
follows that $P$ is valid on every replica, thus ensuring that
assumptions made at a generation event is also valid at the merge
event.

@@

We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
in greater detail. Both withdraw transactions are run at {\sc si}
isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
\underE{\C{SI(Wd2)}}$. We assume an {\sc sc} store, such as a
conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
the maximul visibility weakening (Def.~\ref{def:maxvis}) of {\sc sc}
w.r.t. $\I_c$, as described in \S\ref{sec:store-consistency}. In
concrete terms, this means that an effect added to $\E.\A$ by the
current thread or a concurrent thread immediately becomes visible to
the subsequent operations unless we produce a witness that such
visibility violates $\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note
that the language of our specifications is a decidable subset of the
first-order logic. In particular, our specifications and
rely-guarantee rules are carefully crafted to avoid transitive
closures and quantification over infinite domains. The domain of
assertions is often linear arithmetic, which is also decidable.
Consequently, the proof obligations generated by the application of
our rules can often be discharged via an SMT solver. We nonetheless
give full details in the following.

@@

Decorated parts of \C{Wd1} relevant to the discussion are shown in
Fig.~\ref{fig:wd1-decorated}. \C{Wd2} is similar and not shown.
Temporary name \C{v3} has been introduced to capture the result of the
RHS expression \C{B-a1}.  Assertions implicitly refer to the current
execution ($\E$), just as Hoare triples implicitly refer to the
current state. The context for propositions the value computation of
shared variables is also the implicit, i.e., we write $\psi$ instead
of $\underE{\psi}$ and $\C{B = k}$ instead of $\C{\interp{\E.\A}(B) =
k}$. The proposition \C{k $\ge$ a1+a2} remains an invariant, hence
elided. The pre-condition $P_1$ of \C{Wd1} accounts for the
possibility of \C{Wd2} committing before \C{Wd1}, writing \C{k-a2} to
\C{B}. Since neither \C{Wd1} nor \C{Wd2} are committed at the
beginning, the pre-condition $P$ of the program is extended with
$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
$P_1$ follows. The post-condition $Q_1$ of \C{Wd1} asserts different
values for \C{B} depending on whether or not \C{Wd2} has already
committed.  It also asserts that \C{Wd1} has committed. The
post-condition $Q_2$ of \C{Wd2} (not shown) is similar, but with the
roles of \C{Wd1} and \C{Wd2} exchanged. The post-condition $Q$ follows
from $Q_1 \wedge Q_2$, hence valid. It remains to show that \C{Wd1}
satisfies its specification (\C{Wd2} must also satisfy its spec, but
the proof is similar).

>>>>

First, we observe that the consistency level of a data store can be
captured by store-specific consistency constraints, along with
transaction-specific isolation constraints, via the trace invariant
$\I$. In particular, we can split $\I$ into two components: (1).
$\I_s$, the store-specific invariant, and (2). $\I_c$, the
program-specific (or, client-specific) invariant, to capture
consistency and isolation constraints\footnote{$\I_c$ captures the
combination of transaction-specific isolation constraints.},
respectively.  $\I$ is now a conjunction: $\I \,=\,
\lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often simplified to $\I \,=\,
\I_s\wedge\I_c$). While the program-specific trace invariant ($\I_c$)
is independent of the underlying data store, the store specific
invariant ($\I_s$) changes from store to store depending on the
consistency level.  For an eventually consistent store, $\I_s$ is
simply \emph{true}. Stronger stores, such as those that support strong
consistency, have non-trivial definitions for $\I_s$.

@@

Notably, our view of replication does not explicitly involve replicas.
Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
weakly-consistent replication with a conventional replica-based model.
Under our model, the notion of a replica is subsumed by the concept of
visibility; a replica is defined by the subset ($S$) of global state
($\E.\A$) that an operation witnesses. Constraints over replica states
therefore manifest as constraints over the visibility relation.  For
example, instead of requiring the store to be causally consistent, an
operation can witness a causally consistent subset of the state; such
demands can be made via the trace invariant $\I$. For a pre-condition
($P$) of the operation to be useful, it has to be an assertion over
every causally consistent subset of the global state.  Since any
replica that eventually executes the operation has to expose one such
subset ($S$), the pre-condition is guaranteed to hold regardless of
the replica. There is however one problem with this explanation - by
considering subsets of just one global state, it ignores the fact that
the global state (hence, the replica states) change during the
execution of the operation. To account for such changes, we might
choose to distinguish between an effect generation event at one
replica $r_1$ and an effect merge event at replica $r_2$, requiring
that \emph{non-conflicting} operations execute between these two
events at $r_2$, and that they preserve certain
invariants~\cite{gotsmanpopl16}.  Instead, our framework folds all
such machinery into a stability condition predicated on $\I$
(\S\ref{sec:rely-guarantee}).  Since any change to the global state
during the execution of the operation is an interference, and $P$ is
required to be stable with respect to any such interference, it
follows that $P$ is valid on every replica, thus ensuring that
assumptions made at a generation event is also valid at the merge
event.

@@

We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
in greater detail. Both withdraw transactions are run at {\sc si}
isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
\underE{\C{SI(Wd2)}}$. We assume an {\sc sc} store, such as a
conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
the maximul visibility weakening of {\sc sc} w.r.t. $\I_c$, as
described in \S\ref{sec:store-consistency}. In concrete terms, this
means that an effect added to $\E.\A$ by the current thread or a
concurrent thread immediately becomes visible to the subsequent
operations unless we produce a witness that such visibility violates
$\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note that the language of
our specifications is a decidable subset of the first-order logic. In
particular, our specifications and rely-guarantee rules are carefully
crafted to avoid generating proof obligations with transitive closures
and quantification over infinite domains. The domain of assertions is
often linear arithmetic, which is also decidable. Consequently, the
generated proof obligations can often be discharged via an SMT solver.
We nonetheless give full details in the following.

@@

Decorated parts of \C{Wd1} relevant to the discussion are shown in
Fig.~\ref{fig:wd1-decorated}. \C{Wd2} is similar and not shown.
Temporary name \C{v3} has been introduced to capture the result of the
RHS expression \C{B-a1}.  Assertions implicitly refer to the current
execution ($\E$), just as Hoare triples implicitly refer to the
current state. The context for propositions and the state for shared 
variables is also the implicit, i.e., we write $\psi$ instead
of $\underE{\psi}$ and $\C{B = k}$ instead of $\C{\interp{\E.\A}(B) =
k}$. The proposition \C{k $\ge$ a1+a2} remains an invariant, hence
elided. The pre-condition $P_1$ of \C{Wd1} accounts for the
possibility of \C{Wd2} committing before \C{Wd1}, writing \C{k-a2} to
\C{B}. Since neither \C{Wd1} nor \C{Wd2} are committed at the
beginning, the pre-condition $P$ of the program is extended with
$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
$P_1$ follows. The post-condition $Q_1$ of \C{Wd1} asserts different
values for \C{B} depending on whether or not \C{Wd2} has already
committed.  It also asserts that \C{Wd1} has committed. The
post-condition $Q_2$ of \C{Wd2} (not shown) is similar, but with the
roles of \C{Wd1} and \C{Wd2} exchanged. The post-condition $Q$ follows
from $Q_1 \wedge Q_2$, hence valid. It remains to show that \C{Wd1}
satisfies its specification (\C{Wd2} must also satisfy its spec, but
the proof is similar).
